Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-13 talk/pepm2011/paper.tex
fix some typos, add few XXXs, improve sentences here and there

\newcommand\cfbolz[1]{\nb{CFB}{#1}}
\newcommand\arigo[1]{\nb{AR}{#1}}
\newcommand\fijal[1]{\nb{FIJAL}{#1}}
\newcommand\david[1]{\nb{DAVID}{#1}}
\newcommand\reva[1]{\nb{Reviewer 1}{#1}}
\newcommand\revb[1]{\nb{Reviewer 2}{#1}}
\newcommand\revc[1]{\nb{Reviewer 3}{#1}}
\newcommand{\commentout}[1]{}

@@

A recently popular approach to implementing just-in-time compilers for dynamic
languages is that of a tracing JIT. A tracing JIT works by observing the running
program and recording linear execution traces, which are then turned into
machine code. One reason for the popularity of tracing JITs is their relative
simplicity. They can often be added to an interpreter and a lot of the
infrastructure of an interpreter can be reused. They give some important
optimizations like inlining and constant-folding for free. A tracing JIT always
produces linear pieces of code, which makes many optimizations that are usually
hard in a compiler simpler, such as register allocation.

@@

The feature that makes PyPy more than a compiler with a runtime system is it's
support for automated JIT compiler generation \cite{bolz_tracing_2009}. During
the translation to C, PyPy's tools can generate a just-in-time compiler for the
language that the interpreter is implementing. This process is mostly
automatic; it only needs to be guided by the language implementer by a small number of
source-code hints. Mostly-automatically generating a JIT compiler has many advantages
over writing one manually, which is an error-prone and tedious process.
By construction, the generated JIT has the same semantics as the interpreter.
Many optimizations can benefit all languages implemented as an interpreter in RPython.
The JIT that is produced by PyPy's JIT generator is a \emph{tracing JIT
compiler}, a concept which we now explain in more details.

@@

The core idea of tracing JITs is to focus the optimization effort of the JIT
compiler on the hot paths of the core loops of the program and to just use an
interpreter for the less commonly executed parts. VMs that use a tracing JIT are
thus mixed-mode execution environments, they contain both an interpreter and a
JIT compiler. By default the interpreter is used to execute the program, doing
some light-weight profiling at the same time. This profiling is used to identify
the hot loops of the program. If a hot loop is found in that way, the
interpreter enters a special \emph{tracing mode}. In this tracing mode, the
interpreter records all operations that it is executing while running one
iteration of the hot loop. This history of executed operations of one loop is
called a \emph{trace}. Because the trace corresponds to one iteration of a loop,
it always ends with a jump to its own beginning. The trace also contains all
operations that are performed in functions that were called in the loop, thus a
tracing JIT automatically performs inlining.

@@

If one specific guard fails often enough, the tracing JIT will generate a new
trace that starts exactly at the position of the failing guard. The existing
assembler is patched to jump to the new trace when the guard fails
\cite{andreas_gal_incremental_2006}.

@@

The loop iterates \texttt{y} times, and computes something in the process.
Simply running this function is slow, because there are lots of virtual method
calls inside the loop, one for each \texttt{is\_positive} and even two for each
call to \texttt{add}. These method calls need to check the type of the involved
objects repeatedly and redundantly. In addition, a lot of objects are created
when executing that loop, many of these objects do not survive for very long.
The actual computation that is performed by \texttt{f} is simply a number of
float or integer additions.

\begin{figure}
\texttt{
\begin{tabular}{l} 
\# arguments to the trace: $p_{0}$, $p_{1}$ \\
\# inside f: res.add(y) \\
guard\_class($p_{1}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{2}$ = get($p_{1}$, intval) \\
~~~~guard\_class($p_{0}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{3}$ = get($p_{0}$, intval) \\
~~~~~~~~$i_{4}$ = int\_add($i_{2}$, $i_{3}$) \\
~~~~~~~~$p_{5}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{5}$, intval, $i_{4}$) \\
\# inside f: BoxedInteger(-100)  \\
$p_{6}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~set($p_{6}$, intval, -100) \\
~\\
\# inside f: .add(BoxedInteger(-100)) \\
guard\_class($p_{5}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{7}$ = get($p_{5}$, intval) \\
~~~~guard\_class($p_{6}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{8}$ = get($p_{6}$, intval) \\
~~~~~~~~$i_{9}$ = int\_add($i_{7}$, $i_{8}$) \\
~~~~~~~~$p_{10}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{10}$, intval, $i_{9}$) \\
~\\
\# inside f: BoxedInteger(-1) \\
$p_{11}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~set($p_{11}$, intval, -1) \\
~\\
\# inside f: y.add(BoxedInteger(-1)) \\
guard\_class($p_{0}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{12}$ = get($p_{0}$, intval) \\
~~~~guard\_class($p_{11}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{13}$ = get($p_{11}$, intval) \\
~~~~~~~~$i_{14}$ = int\_add($i_{12}$, $i_{13}$) \\
~~~~~~~~$p_{15}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{15}$, intval, $i_{14}$) \\
~\\
\# inside f: y.is\_positive() \\
guard\_class($p_{15}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.is\_positive \\
~~~~$i_{16}$ = get($p_{15}$, intval) \\
~~~~$i_{17}$ = int\_gt($i_{16}$, 0) \\
\# inside f \\
guard\_true($i_{17}$) \\
jump($p_{15}$, $p_{10}$) \\
\end{tabular}
}
\caption{Unoptimized Trace for the Simple Object Model}
\label{fig:unopt-trace}
\end{figure}

@@

The partial evaluation works by walking the trace from beginning to end.
Whenever a \texttt{new} operation is seen, the operation is removed and a static
object is constructed and associated with the variable that would have stored
the result of \texttt{new}. The static object describes the shape of the
original object, \eg where the values that would be stored in the fields of the
allocated object come from, as well as the type of the object. Whenever the
optimizer sees a \texttt{set} that writes into such an object, that shape
description is updated and the operation can be removed, which means that the
operation was done at partial evaluation time. When the optimizer encounters a
\texttt{get} from such an object, the result is read from the shape
description, and the operation is also removed. Equivalently, a
\texttt{guard\_class} on a variable that has a shape description can be removed
as well, because the shape description stores the type.

@@

The static object associated with $p_{5}$ would know that it is a
\texttt{BoxedInteger} whose \texttt{intval} field contains $i_{4}$; the
one associated with $p_{6}$ would know that it is a \texttt{BoxedInteger}
whose \texttt{intval} field contains the constant -100.

@@

This is what happens at the end of the trace in Figure~\ref{fig:unopt-trace}, when the \texttt{jump} operation
is hit. The arguments of the jump are at this point static objects. Before the
jump is emitted, they are \emph{lifted}. This means that the optimizer produces code
that allocates a new object of the right type and sets its fields to the field
values that the static object has (if the static object points to other static
objects, those need to be lifted as well) This means that instead of the jump,
the following operations are emitted:

@@

Note how the operations for creating these two instances have been moved down the
trace. It looks like for these operations we actually didn't win much, because
the objects are still allocated at the end. However, the optimization was still
worthwhile even in this case, because some operations that have been performed
on the lifted static objects have been removed (some \texttt{get} operations
and \texttt{guard\_class} operations).

@@

If the argument $v$ of a \texttt{get} operation is mapped to something in the static
heap, the get can be performed at optimization time. Otherwise, the \texttt{get}
operation needs to be residualized.

@@

Furthermore there are the following benchmarks:
\begin{itemize}
    \item \textbf{crypto\_pyaes}: AES implementation.
    \item \textbf{django}: The templating engine of the Django web
    framework\footnote{\texttt{http://www.djangoproject.com/}}.
    \item \textbf{go}: A Monte-Carlo Go
    AI\footnote{\texttt{http://shed-skin.blogspot.com/2009/07/ disco-elegant-python-go-player.html}}.
    \item \textbf{html5lib}: An HTML5 parser.
    \item \textbf{pyflate-fast}: A BZ2 decoder.
    \item \textbf{raytrace-simple}: A ray tracer.
    \item \textbf{richards}: The Richards benchmark.
    \item \textbf{spambayes}: A Bayesian spam filter\footnote{\texttt{http://spambayes.sourceforge.net/}}.
    \item \textbf{telco}: A Python version of the Telco decimal.
    benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}},
    using a pure Python decimal floating point implementation.
    \item \textbf{twisted\_names}: A DNS server benchmark using the Twisted networking
    framework\footnote{\texttt{http://twistedmatrix.com/}}.
\end{itemize}

We evaluate the allocation removal algorithm along two lines: First we want to
know how many allocations could be optimized away. On the other hand, we want
to know how much the run times of the benchmarks is improved.

For the former we counted the occurring operations in all generated traces
before and after the optimization phase for all benchmarks. The results can be
seen in Figure~\ref{fig:numops}. The optimization removes as many as XXX and as
little as XXX percent of allocation operations in the benchmarks. All benchmarks
taken together, the optimization removes XXX percent of allocation operations.

@@

In addition to the count of operations we also performed time measurements. The
machine the benchmarks were run on is XXX. We compared the performance of
various Python implementations on the benchmarks. As a baseline, we used the
standard Python implementation in C, called
CPython\footnote{\texttt{http://python.org}}, which uses a bytecode-based
interpreter. Furthermore we compared against Psyco \cite{rigo_representation-based_2004}, an extension to
CPython which is a just-in-time compiler that produces machine code at run-time.
It is not based on traces. Of PyPy's Python interpreter we used three versions,
one without a JIT, one including the JIT but not using the allocation removal
optimization, and one using the allocation removal optimizations.

>>>>

\newcommand\cfbolz[1]{\nb{CFB}{#1}}
\newcommand\arigo[1]{\nb{AR}{#1}}
\newcommand\fijal[1]{\nb{FIJAL}{#1}}
\newcommand\david[1]{\nb{DAVID}{#1}}
\newcommand\anto[1]{\nb{ANTO}{#1}}
\newcommand\reva[1]{\nb{Reviewer 1}{#1}}
\newcommand\revb[1]{\nb{Reviewer 2}{#1}}
\newcommand\revc[1]{\nb{Reviewer 3}{#1}}
\newcommand{\commentout}[1]{}

@@

A recently popular approach to implementing just-in-time compilers for dynamic
languages is that of a tracing JIT. A tracing JIT works by observing the running
program and recording its hot parts into linear execution traces, which are then turned into
machine code. One reason for the popularity of tracing JITs is their relative
simplicity. They can often be added to an interpreter and a lot of the
infrastructure of an interpreter can be reused. They give some important
optimizations like inlining and constant-folding for free. A tracing JIT always
produces linear pieces of code, which simplifies many optimizations that are usually
hard in a compiler, such as register allocation.

@@

The feature that makes PyPy more than a compiler with a runtime system is its
support for automated JIT compiler generation \cite{bolz_tracing_2009}. During
the translation to C, PyPy's tools can generate a just-in-time compiler for the
language that the interpreter is implementing. This process is mostly
automatic; it only needs to be guided by the language implementer by a small number of
source-code hints. Mostly-automatically generating a JIT compiler has many advantages
over writing one manually, which is an error-prone and tedious process.
By construction, the generated JIT has the same semantics as the interpreter.
Many optimizations can benefit all languages implemented as an interpreter in RPython.

Moreover, thanks to the internal design of the JIT generator, it is very easy
to add new \emph{backends} for producing the actual machine code, in addition
to the original backend for the Intel \emph{x86} architecture.  Examples of
additional JIT backends are the one for Intel \emph{x86-64} and an
experimental one for the CLI .NET Virtual Machine \cite{cuni_python_cli_2010}.
The JIT that is produced by PyPy's JIT generator is a \emph{tracing JIT
compiler}, a concept which we now explain in more details.

@@

The core idea of tracing JITs is to focus the optimization effort of the JIT
compiler on the hot paths of the core loops of the program and to just use an
interpreter for the less commonly executed parts. VMs that use a tracing JIT are
thus mixed-mode execution environments, they contain both an interpreter and a
JIT compiler. By default the interpreter is used to execute the program, doing
some light-weight profiling at the same time. This profiling is used to identify
the hot loops of the program. If a hot loop is found in that way, the
interpreter enters a special \emph{tracing mode}. In this tracing mode, the
interpreter records all operations that it is executing while running one
iteration of the hot loop. This history of executed operations of one loop is
called a \emph{trace}. Because the trace corresponds to one iteration of a loop,
it always ends with a jump to its own beginning \anto{this is not true: what if we trace the last iteration?}. The trace also contains all
operations that are performed in functions that were called in the loop, thus a
tracing JIT automatically performs inlining.

@@

If one specific guard fails often enough, the tracing JIT will generate a new
trace that starts exactly at the position of the failing guard. The existing
assembler is patched to jump to the new trace when the guard fails
\cite{andreas_gal_incremental_2006}.  This approach guarantees that all the
hot paths in the program will eventually be traced and compiled into efficient
code.

@@

The loop iterates \texttt{y} times, and computes something in the process.
Simply running this function is slow, because there are lots of virtual method
calls inside the loop, one for each \texttt{is\_positive} and even two for each
call to \texttt{add}. These method calls need to check the type of the involved
objects repeatedly and redundantly. In addition, a lot of objects are created
when executing that loop, many of these objects do not survive for very long.
The actual computation that is performed by \texttt{f} is simply a sequence of
float or integer additions.

\begin{figure}
\texttt{
\begin{tabular}{l} 
\# XXX: maybe we should specify that $p_{0}$, $p_{1}$ corresponds to y and res
\# arguments to the trace: $p_{0}$, $p_{1}$ \\
\# inside f: res.add(y) \\
guard\_class($p_{1}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{2}$ = get($p_{1}$, intval) \\
~~~~guard\_class($p_{0}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{3}$ = get($p_{0}$, intval) \\
~~~~~~~~$i_{4}$ = int\_add($i_{2}$, $i_{3}$) \\
~~~~~~~~$p_{5}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{5}$, intval, $i_{4}$) \\
\# inside f: BoxedInteger(-100)  \\
$p_{6}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~set($p_{6}$, intval, -100) \\
~\\
\# inside f: .add(BoxedInteger(-100)) \\
guard\_class($p_{5}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{7}$ = get($p_{5}$, intval) \\
~~~~guard\_class($p_{6}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{8}$ = get($p_{6}$, intval) \\
~~~~~~~~$i_{9}$ = int\_add($i_{7}$, $i_{8}$) \\
~~~~~~~~$p_{10}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{10}$, intval, $i_{9}$) \\
~\\
\# inside f: BoxedInteger(-1) \\
$p_{11}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~set($p_{11}$, intval, -1) \\
~\\
\# inside f: y.add(BoxedInteger(-1)) \\
guard\_class($p_{0}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{12}$ = get($p_{0}$, intval) \\
~~~~guard\_class($p_{11}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{13}$ = get($p_{11}$, intval) \\
~~~~~~~~$i_{14}$ = int\_add($i_{12}$, $i_{13}$) \\
~~~~~~~~$p_{15}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{15}$, intval, $i_{14}$) \\
~\\
\# inside f: y.is\_positive() \\
guard\_class($p_{15}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.is\_positive \\
~~~~$i_{16}$ = get($p_{15}$, intval) \\
~~~~$i_{17}$ = int\_gt($i_{16}$, 0) \\
\# inside f \\
guard\_true($i_{17}$) \\
jump($p_{15}$, $p_{10}$) \\
\end{tabular}
}
\caption{Unoptimized Trace for the Simple Object Model}
\label{fig:unopt-trace}
\end{figure}

@@

The partial evaluation works by walking the trace from beginning to end.
Whenever a \texttt{new} operation is seen, the operation is removed and a static
object is constructed and associated with the variable that would have stored
the result of \texttt{new}. The static object describes the shape of the
original object, \eg where the values that would be stored in the fields of the
allocated object come from, as well as the type of the object. Whenever the
optimizer sees a \texttt{set} that writes into such an object, that shape
description is updated and the operation can be removed, which means that the
operation was done at partial evaluation time. When the optimizer encounters a
\texttt{get} from such an object, the result is read from the shape
description, and the operation is also removed. Equivalently, a
\texttt{guard\_class} on a variable that has a shape description can be removed
as well, because the shape description stores the type and thus the result of 
the guard is statically known.

@@

The static object associated with $p_{5}$ would know that it is a
\texttt{BoxedInteger} whose \texttt{intval} field contains $i_{4}$; the
one associated with $p_{6}$ would know that it is a \texttt{BoxedInteger}
whose \texttt{intval} field contains the constant -100.
\anto{this works only because we use SSI and thus the value of $i_{4}$ never
changes. However, SSI is not explained anywhere in the paper}

@@

This is what happens at the end of the trace in Figure~\ref{fig:unopt-trace}, when the \texttt{jump} operation
is hit. The arguments of the jump are at this point static objects. Before the
jump is emitted, they are \emph{lifted}. This means that the optimizer produces code
that allocates a new object of the right type and sets its fields to the field
values that the static object has (if the static object points to other static
objects, those need to be lifted as well, recursively) This means that instead of the jump,
the following operations are emitted:

@@

Note how the operations for creating these two instances have been moved down the
trace. It looks like for these operations we actually didn't win much, because
the objects are still allocated at the end. However, the optimization was still
worthwhile even in this case, because some operations that have been performed
on the lifted static objects have been removed (some \texttt{get} operations
and \texttt{guard\_class} operations).  Moreover, in real life example usually
the loops are more complex and contain more objects of type 1, thus this
technique is more effective.

@@

If the argument $v$ of a \texttt{get} operation is mapped to something in the static
heap, the \texttt{get} can be performed at optimization time. Otherwise, the \texttt{get}
operation needs to be residualized.

@@

Furthermore there are the following benchmarks:
\begin{itemize}
    \item \textbf{crypto\_pyaes}: AES implementation.
    \item \textbf{django}: The templating engine of the Django web
    framework\footnote{\texttt{http://www.djangoproject.com/}}.
    \item \textbf{go}: A Monte-Carlo Go
    AI\footnote{\texttt{http://shed-skin.blogspot.com/2009/07/ disco-elegant-python-go-player.html}}.
    \item \textbf{html5lib}: An HTML5 parser.
    \item \textbf{pyflate-fast}: A BZ2 decoder.
    \item \textbf{raytrace-simple}: A ray tracer.
    \item \textbf{richards}: The Richards benchmark.
    \item \textbf{spambayes}: A Bayesian spam filter\footnote{\texttt{http://spambayes.sourceforge.net/}}.
    \item \textbf{telco}: A Python version of the Telco decimal
    benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}},
    using a pure Python decimal floating point implementation.
    \item \textbf{twisted\_names}: A DNS server benchmark using the Twisted networking
    framework\footnote{\texttt{http://twistedmatrix.com/}}.
\end{itemize}

We evaluate the allocation removal algorithm along two lines: first we want to
know how many allocations could be optimized away. On the other hand, we want
to know how much the run times of the benchmarks is improved.

For the former we counted the occurring operations in all generated traces
before and after the optimization phase for all benchmarks. The results can be
seen in Figure~\ref{fig:numops}. The optimization removes as many as XXX and as
little as XXX percent of allocation operations in the benchmarks. All benchmarks
taken together, the optimization removes XXX percent of allocation operations.
\anto{Actually, we can only know how many operations we removed from the traces, not from the actual execution}

@@

In addition to the count of operations we also performed time measurements. The
machine the benchmarks were run on is XXX. We compared the performance of
various Python implementations on the benchmarks. As a baseline, we used the
standard Python implementation in C, called
CPython\footnote{\texttt{http://python.org}}, which uses a bytecode-based
interpreter. Furthermore we compared against Psyco \cite{rigo_representation-based_2004}, an extension to
CPython which is a just-in-time compiler that produces machine code at run-time.
It is not based on traces. Finally, we used three versions of PyPy's Python interpreter:
one without a JIT, one including the JIT but not using the allocation removal
optimization, and one using the allocation removal optimizations.
